Classless Java: Tuning Java Interfaces

### 2015-12-09 paper/sections/Formalization.tex
Done Spacing
comments removed
small fixes

\noindent
What happens if a new interface \Q@M@ extends one \Q@A@${}_i$ and one
\Q@B@${}_j$?\\*
% Nine representative cases are shown next:\\*
\noindent
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{M extends} & \textbf{A1}                  & \textbf{A2} & \textbf{A3} \\ \hline
\textbf{B1}        & conservative error                        & conflict error      & conservative error       \\ \hline
\textbf{B2}        & both abstract, accepted                        & conservative error       & both abstract, accepted       \\ \hline
\textbf{B3}        & \textbf{conservative error} &conflict  error       & conservative error      \\ \hline
\end{tabular}
\\*
%We try to classify the results in the table:
\begin{itemize}
\item \textbf{conflict error} happens when the methods from both interfaces are implemented, which is also an error in most trait models.
\item \textbf{both abstract, accepted} happens when the methods from both interfaces are abstract, which is also considered correct in all
  trait models.
\item \textbf{conservative error} happens when only one method is implemented
  (leaving another one abstract), which is different from what we would expect in
  a trait model, but is coherent with the conservative idea that a method
  defined in an interface should not silently satisfy a method in another one.
\end{itemize}

@@

\begin{figure}[t]
\begin{grammar}
\production{
\e
}{
  \x\mid\MCall\e\m\es\mid\MCall{\C}\m\es\mid\MCall{\C\QM{.super}}\m\es\mid\x\QM=\e\QM;\e'\mid\obj
  }{expressions}\\
\production{
\obj
}{
\QM{new}\ \C\oR\cR\oC\fields\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
  }{object creation}\\
\production{\field}{\T\ \f \QM= \x\QM;}{field declaration}\\
\production{
\II
}{
 \ann\ \QM{interface}\ \C\ \QM{extends}\ \Cs\ \oC \methods\ \cC
  }{interface declaration}\\
\production{
\method
}{
 \QM{static}\ \mh\ \oC\QM{return}\ \e\QM{;}\!\cC
\mid
\QM{default}\ \mh\ \oC\QM{return}\ \e\QM{;}\cC
\mid
\mh\QM{;}
  }{method declaration}\\
\production{
\mh
}{
 \T_0\ \m\ \oR\T_1\ \x_1\ldots\T_n\ \x_n\cR
  }{method header}\\
\production{
\ann
}{
  \mixinAnn|\emptyset
  }{annotations}\\
\production{\Gamma}{
\x_1{:}\C_1\ldots\x_n{:}\C_n
}{environment}
\end{grammar}
\caption{Grammar of Classless Java}
\label{Grammar}
\end{figure}

@@

\begin{figure}[t]
$
\begin{array}{l}

@@

\yanlin{I think the explaination for needed might be hard to understand (for
  reviewer). need revising.}
\paragraph{\tops{}} recovers from the interface table only the ``needed'' methods, that is,
the non-static ones that are not reachable by another, less specific superinterface.
Since the second parameter of \tops{} is a set, we can choose an arbitrary element to be $\C_0$.
In the definition we denote by $\Aux{originalMethod}(\m,\C)=\method$ the non-static method called $\m$ defined directly in $\C$.
%\hl{Another two functions $\super(\C_0)$ and $\get(\m,\C_0)$ are used for the definition. $\super(\C_0)$ collects the
%super interfaces of $\C_0$ recursively, and $\get(\m,\C_0)$ returns the non-static method declared in $\C_0$ with name $\m$, or $\emptyset$ otherwise.
%Finally, $\tops(\m,\C_0)$ returns the topmost methods $m$ from the super interfaces.}
Formally:

@@

\yanlin{added a colon}
\noindent The definition $\Aux{mostSpecific}$ returns the most
specific method whose type is the subtype of all the
others.  Since method subtyping is a partial ordering,
$\Aux{mostSpecific}$ may not be defined, this in turn forces us to rely on the last clause of $\override$; otherwise
the whole \mBody{} would not be defined for that specific $\m$.
Rule \rn{t-intf} relies on this behaviour.

>>>>

\noindent
What happens if a new interface \Q@M@ extends one \Q@A@${}_i$ and one
\Q@B@${}_j$?\\*
% Nine representative cases are shown next:\\*
\noindent
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{M extends} & \textbf{A1}                  & \textbf{A2} & \textbf{A3} \\ \hline
\textbf{B1}        & conservative error                     & conflict error      & conservative error       \\ \hline
\textbf{B2}        & both abstract (accepted)     & conservative error       & both abstract (accepted)       \\ \hline
\textbf{B3}        & \textbf{conservative error} &conflict  error       & conservative error      \\ \hline
\end{tabular}
\\*
%We try to classify the results in the table:
\begin{itemize}
\item \textbf{conflict error} happens when the methods from both interfaces are implemented, which is also an error in most trait models.
\item \textbf{both abstract (accepted)} happens when the methods from both interfaces are abstract, which is also considered correct in all
  trait models.
\item \textbf{conservative error} happens when only one method is implemented
  (leaving another one abstract), which is different from what we would expect in
  a trait model, but is coherent with the conservative idea that a method
  defined in an interface should not silently satisfy a method in another one.
\end{itemize}

@@

\begin{figure}[t]
\saveSpaceFig
\begin{grammar}
\production{
\e
}{
  \x\mid\MCall\e\m\es\mid\MCall{\C}\m\es\mid\MCall{\C\QM{.super}}\m\es\mid\x\QM=\e\QM;\e'\mid\obj
  }{expressions}\\
\production{
\obj
}{
\QM{new}\ \C\oR\cR\oC\fields\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
  }{object creation}\\
\production{\field}{\T\ \f \QM= \x\QM;}{field declaration}\\
\production{
\II
}{
 \ann\ \QM{interface}\ \C\ \QM{extends}\ \Cs\ \oC \methods\ \cC
  }{interface declaration}\\
\production{
\method
}{
 \QM{static}\ \mh\ \oC\QM{return}\ \e\QM{;}\!\cC
\mid
\QM{default}\ \mh\ \oC\QM{return}\ \e\QM{;}\cC
\mid
\mh\QM{;}
  }{method declaration}\\
\production{
\mh
}{
 \T_0\ \m\ \oR\T_1\ \x_1\ldots\T_n\ \x_n\cR
  }{method header}\\
\production{
\ann
}{
  \mixinAnn|\emptyset
  }{annotations}\\
\production{\Gamma}{
\x_1{:}\C_1\ldots\x_n{:}\C_n
}{environment}
\end{grammar}
\caption{Grammar of Classless Java}
\label{Grammar}
\end{figure}

@@

\begin{figure}[t]
\saveSpaceFig
$
\begin{array}{l}

@@

\paragraph{\tops{}} recovers from the interface table only the ``needed'' methods, that is,
the non-static ones that are not reachable by another, less specific superinterface.
Since the second parameter of \tops{} is a set, we can choose an arbitrary element to be $\C_0$.
In the definition we denote by $\Aux{originalMethod}(\m,\C)=\method$ the non-static method called $\m$ defined directly in $\C$.
%\hl{Another two functions $\super(\C_0)$ and $\get(\m,\C_0)$ are used for the definition. $\super(\C_0)$ collects the
%super interfaces of $\C_0$ recursively, and $\get(\m,\C_0)$ returns the non-static method declared in $\C_0$ with name $\m$, or $\emptyset$ otherwise.
%Finally, $\tops(\m,\C_0)$ returns the topmost methods $m$ from the super interfaces.}
Formally:

@@

\noindent The definition $\Aux{mostSpecific}$ returns the most
specific method whose type is the subtype of all the
others.  Since method subtyping is a partial ordering,
$\Aux{mostSpecific}$ may not be defined, this in turn forces us to rely on the last clause of $\override$; otherwise
the whole \mBody{} would not be defined for that specific $\m$.
Rule \rn{t-intf} relies on this behaviour.

### 2015-12-09 paper/sections/Introduction.tex
small fixes

Java 8 introduced \emph{default methods}, allowing interfaces to have
method implementations. The main motivation behind the introduction of
default methods in Java 8 is \emph{interface evolution}. That is, to
allow interfaces to be extended over time while preserving backwards
compatibility. It soon became clear that default methods could also be
used to emulate something similar to \emph{traits}~\cite{scharli03traits}. The
original notion of traits by Scharli et al. prescribes, among other
things, that: 1) a trait provides a set of methods that implement
behaviour; and 2) a trait does not specify any state variables, so the
methods provided by traits do not access state variables
directly. Java 8 interfaces follow similar principles too. Indeed, a
detailed description of how to emulate trait-oriented programming in
Java 8 can be found in the work by Bono et al.~\cite{bono14}. The Java 8
team designing default methods, was also fully aware of that secondary
use of interfaces, but it was not their objective to model traits:
``The key goal of adding default methods to Java was "interface
evolution", not "poor man's traits"''~\cite{goetz13default}. As a result
they were happy to support the secondary use of interfaces with
default methods as long as it did not make the implementation and
language more complex.

@@

\item {\em Interfaces have no constructors.} As a result classes are 
still required to create objects, leading to substantial boilerplate 
code for initialization.

@@

Object interfaces do not require changes to the Java runtime or compiler, 
and they also do not introduce any new syntax. All three features of object interfaces are
achieved by reinterpreting existing Java syntax, and are translated
into regular Java code without loss of type-safety. Since no new
syntax is introduced, it would be incorrect to call object interfaces
a language extension or syntactic sugar. So we use the term
\emph{language tuning} instead. Language tuning sits in between a
lightweight language extension and a glorified library. Language
tuning can offer many features usually implemented by a real language
extension, but because it does not modify the language syntax
pre-existing tools can work transparently on the tuned language.  To
exploit the full benefits of language tuning, our prototype
implementation of object interfaces uses Java annotations to do AST
rewriting, allowing existing Java tools (such as IDEs) to work
out-of-the-box with our implementation. As a result we could
experiment object interfaces with several interesting Java programs,
and conduct various case studies.

To formalize object interfaces, we propose Classless Java (CJ): a
FeatherweightJava-style~\cite{Igarashi01FJ} calculus, which captures the essence
of interfaces with default methods. The semantics of object interfaces
is given as a syntax-directed translation from CJ to itself. In the
resulting CJ code all object interfaces are translated into regular CJ
(and Java) interfaces with default methods. The translation is proved
to be type-safe, ensuring that the translation does not
introduce type-errors in client code. 
%CJ's usefulness goes beyond serving as a
%calculus to formalize object interfaces. During the development
%process of CJ, we encountered a bug in the implementation of default
%methods for the Eclipse Compiler for Java (ECJ). For the program revealing the 
%bug, ECJ behaves differently from both our formalization and Oracle's 
%Java compiler.

@@

\item {\bf Implementation and Case Studies:} We have a prototype
  implementation of object interfaces, using Java
  annotations and AST rewriting. Moreover the usefulness of object interfaces is
  illustrated through various examples and case studies.

>>>>

Java 8 introduced \emph{default methods}, allowing interfaces to have
method implementations. The main motivation behind the introduction of
default methods in Java 8 is \emph{interface evolution}. That is, to
allow interfaces to be extended over time while preserving backwards
compatibility. It soon became clear that default methods could also be
used to emulate something similar to \emph{traits}~\cite{scharli03traits}. The
original notion of traits by Scharli et al. prescribes, among other
things, that: 1) a trait provides a set of methods that implement
behaviour; and 2) a trait does not specify any state variables, so the
methods provided by traits do not access state variables
directly. Java 8 interfaces follow similar principles too. Indeed, a
detailed description of how to emulate trait-oriented programming in
Java 8 can be found in the work by Bono et al.~\cite{bono14}. The Java 8
team designing default methods, was also fully aware of that secondary
use of interfaces, but it was not their objective to model traits:
``The key goal of adding default methods to Java was "interface
evolution", not "poor man's traits"''~\cite{goetz13default}. As a result, 
they were happy to support the secondary use of interfaces with
default methods as long as it did not make the implementation and
language more complex.

@@

\item {\em Interfaces have no constructors.} As a result, classes are 
still required to create objects, leading to substantial boilerplate 
code for initialization.

@@

Object interfaces do not require changes to the Java runtime or compiler, 
and they also do not introduce any new syntax. All three features of object interfaces are
achieved by reinterpreting existing Java syntax, and are translated
into regular Java code without loss of type-safety. Since no new
syntax is introduced, it would be incorrect to call object interfaces
a language extension or syntactic sugar. So we use the term
\emph{language tuning} instead. Language tuning sits in between a
lightweight language extension and a glorified library. Language
tuning can offer many features usually implemented by a real language
extension, but because it does not modify the language syntax
pre-existing tools can work transparently on the tuned language.  To
exploit the full benefits of language tuning, our prototype
implementation of object interfaces uses Java annotations to do AST
rewriting, allowing existing Java tools (such as IDEs) to work
out-of-the-box with our implementation. As a result, we could
experiment object interfaces with several interesting Java programs,
and conduct various case studies.

To formalize object interfaces, we propose Classless Java (CJ): a
FeatherweightJava-style~\cite{Igarashi01FJ} calculus, which captures the essence
of interfaces with default methods. The semantics of object interfaces
is given as a syntax-directed translation from CJ to itself. In the
resulting CJ code, all object interfaces are translated into regular CJ
(and Java) interfaces with default methods. The translation is proved
to be type-safe, ensuring that the translation does not
introduce type-errors in client code. 
%CJ's usefulness goes beyond serving as a
%calculus to formalize object interfaces. During the development
%process of CJ, we encountered a bug in the implementation of default
%methods for the Eclipse Compiler for Java (ECJ). For the program revealing the 
%bug, ECJ behaves differently from both our formalization and Oracle's 
%Java compiler.

@@

\item {\bf Implementation and Case Studies:} We have a prototype
  implementation of object interfaces, using Java
  annotations and AST rewriting. Moreover, the usefulness of object interfaces is
  illustrated through various examples and case studies.
